# Functions

## Define your very own R function
Many times, we want to re-use a chunk of code. The most efficient way is to wrap these code as a function, clearly define what the input and the output. Functions are fundamental building blocks of R. Most of the times when we run R commands, we are calling and executing functions.  We can easily define our own functions.

For example, we have the following arithmetic function: 

$$f(x)=1.5 x^3+ x^2-2x+1$$

Obviously, we can use the following code to do the computing:
```{r}
x <- 5
1.57345 * x ^ 3 + x ^ 2 - 2 * x + 1
```

This will work, but every time we have to re-write this code. 
So let’s try to define our own function:
```{r}
myf <- function(x) {
       y = 1.57345 * x ^ 3 + x ^ 2 - 2 * x + 1
       return(y)
}
```

Note that “{”  and “}” signify the beginning and end of a block of code.  **“function”** tells R that a function is going to be defined. At the end, the **“return”** statement returns the desired value. 

You can copy and paste the 4 lines of code to R and it defines a function called myf, which you can call by:
```{r}
myf(5)  # or myf(x = 5)
```

As you can see you get the same results when x=5. But now you can use this in many ways. 
```{r}
x <- - 10 : 10  # x now is a vector  with 21 numbers -10, -9, … 10
myf(x) 
plot(x, myf(x))  # see plot on the right.
```

Obviously functions can handle many different calculations beyond arithmetic functions. It can take in one or more inputs and return a list of complex data objects too. 

Optional. Submit solutions of these challenges to get 5 bonus points. 

**Challenge 1**: Write an R function to implement this arithmetic function:  f(x)= √(|x|  )+5x-6. Note |x| means the absolute value of x. Use this function to find f(4.534), and also try to produce a plot like the one above. 

**Challenge 2**: Go over the **R cheat sheet**, find some function that was not covered in class and show some demo. 

