# Data exploring of state dataset

## Read in data and manage data
 
There are 7 little data sets in the state data group. We pick three of them, including state.abb, state.x77, and state.region to form our data file "sta". The detail information about the variables *in the three data sets are listed here.

**state.abb**:
character vector of 2-letter abbreviations for the state names.

**state.x77**:
matrix with 50 rows and 8 columns giving the following statistics in the respective columns.

Population:
population estimate as of July 1, 1975

Income:
per capita income (1974)

Illiteracy:
illiteracy (1970, percent of population)

Life Exp:
life expectancy in years (1969-71)

Murder:
murder and non-negligent manslaughter rate per 100,000 population (1976)

HS Grad:
percent high-school graduates (1970)

Frost:
mean number of days with minimum temperature below freezing (1931-1960) in capital or large city

Area:
land area in square miles

**state.region**:
factor giving the region (Northeast, South, North Central, West) that each state belongs to.

We merge these three and got final dataframe "sta" with 10 columns and 50 rows.
```{r}
tem <- cbind(state.abb, state.x77, state.region)
class(tem)
sta <- data.frame(tem)  #change tem from matrix to dataframe
head(sta)
dim(sta)
colnames(sta)[1] <- "State"
colnames(sta)[10] <- "Region"
str(sta)
```

```{r}
sta$Population <- as.numeric(sta$Population)
sta$Income <- as.numeric(sta$Income)
sta$Illiteracy <- as.numeric(sta$Illiteracy)
sta$Life.Exp <- as.numeric(sta$Life.Exp)
sta$Murder <- as.numeric(sta$Murder)
sta$HS.Grad <- as.numeric(sta$HS.Grad)
sta$Frost <- as.numeric(sta$Frost)
sta$Area <- as.numeric(sta$Area)
```

## Analyzing each set of data column seperately






