---
title: "Homework 4"
author: "Rong Fan"
date: "February 3, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{exercise}
Exercise 3.1
Now try multiplying 45.6 by 78.9.
```

```{r}
45.6*78.9
```

```{exercise}
Exercise 3.2
Suppose a vector is definded as x <- c(12, 56, 31, -5, 7). Calculate the mean of all elements in x, assign the mean to y. Squared each element in x and assign the result in a new vector z.
```


```{r}
x <- c(12, 56, 31, -5, 7)
y <- mean(x)
y
```
```{r}
z <- x^2
z
```

```{exercise}
Exercise 3.3
Try to find and run the two functions that sets and returns the current working directory.
```
> setwd("C:/Users/rong.fan/Google Drive/teaching/")

> getwd()

[1] "C:/Users/rong.fan/Google Drive/teaching"

```{exercise}
Exercise 3.4
Try to find and run the function that lists all the files in the current working folder.
```

```{r results = 'hide'}

list.files(path = ".")
```


```{exercise}
Exercise 3.5
Write an R function to implement this arithmetic function: f(x)= |x|+5x-6. Note |x| means the absolute value of x. Use this function to find f(4.534), and also produce a plot like the example.
Let’s define a function to count even numbers in a vector.
```

```{r}
myfun.1 <- function(x){
  y <- abs(x)+5*x-6
  return(y)
}
myfun.1(4.534)
```


```{exercise}
Exercise 3.6
Define a function counting the values that are less than 0 for two vectors x=rnorm(50) and y=rnorm(5000) respectively. Define another function to calcute the proportion that values are less than 0 for x and y respectively. Compare calculated proportions with theoretical proportion 0.5, what conclusions can you make?
```
```{r}
x=rnorm(50)
y=rnorm(5000)
```

```{r}
count.fun <- function(x){
  k <- 0
  for (n in x) { 
    if (n < 0) k <- k+1
  } 
  return(k)
}
```

```{r}
count.fun(x)
```
```{r}
count.fun(y)
```

```{r}
prop.fun <- function(x){
  k <- 0
  for (n in x) { 
    if (n < 0) k <- k+1
  } 
  return(k/length(x))
}
```

```{r}
prop.fun(x)

```
```{r}
prop.fun(y)
```

Conclusion: The x and y are two random sets of numbers following standard normal distribution. The number of values in x, also called a sample size, is 50. The sample size for y is 5000. If a random variable following standard normal distribution, the probability that the random variable less than 0 is $P(X<0)=\Phi(0)=0.5$. Comparing the two proportions, we can see that the proportion of the negative numbers in y is more closer to 0.5 than that of in x. The main difference between x and y is that y has a much larger sample size than x. The result implies that the proportion obtained from larger sample size is more accurate. Even though the conclusion is not always true due to randomness.

```{exercise}
Exercise 3.7 
Compute 1+2+3… +1000 with one line of R code. Hint: examine the example code for sum( ) function in the R help document.
```
```{r}
sum(1:1000)
```

```{exercise}
Exercise 3.8
Using sample selction function randomly select 10 integers from 1 to 100. Create a vector y which satisfies the following conditions: if an selcted integer is an even number, then y returns ‘even’, otherwise y returns ‘odd’.
```

```{r}
x <- sample(1:100,10)
x
y <- ifelse (x %% 2 == 0, 'even', 'odd') 
y
```

```{exercise}
Exercise 3.9 Now compute the average of values in z. Ignore the missing values.
```

```{r}
x <- c(5, 2, 22, 11, 5)
y <- c(5, 11, 8)
z <- match(y, x)
mean(z, na.rm = TRUE)
```

```{exercise}
Exercise 3.10 Create a vector with 21 integers from -10 to 10, and store it in the x variable. Then create a scatterplot of x^2 against x.
```

```{r}
x <- seq(from = -10, to = 10, length = 21)
x
```
```{r}
y <- x^2
plot(x,y)
```



```{exercise}
Exercise 3.11 Does Mickey caught more fishes than Tom and Jerry combined? Write R code to verify this statement using the fishes vector and return a TRUE or FALSE value.
```
```{r}
fishes <- c(7, 3, 9)
fishes[3] > fishes[1]+fishes[2]
```

```{exercise}
Exercise 3.12 Now see if you can set the value for Tom to something other than 5 using the name rather than the index.
```
```{r}
fishes <- c(7, 3, 9)
names(fishes) <- c("Tom", "Jerry", "Mickey")
fishes['Tom'] <- 'Ten'
fishes
```


```{exercise}
Exercise 3.13 Create a vector representing the prices of groceries, bread $2.5, milk $3.1, jam $5.3, beer $9.1. And create a bar plot to represent this information.
```
```{r}
price <- c(2.5, 3.1, 5.3, 9.1)
item <- names(price) <- c('bread', 'milk', 'jam', 'beer')
price
```
```{r}
barplot(price)
```


