---
title: "HW6 Solution"
author: "Rong Fan"
date: "February 9, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{exercise}
Exercise 3.18 Access the second component in list1 by three ways as we introduced above. What is the types of list1["Hobby"] and list1[["Hobby"]] respectively?
```{r}
  list1 <- list(Name = "Tom", Hobby = "Fishing", Num.fish = c(16, 27, 5))
```
  
Solution:
```{r}
list1[[2]]
list1$Hobby
list1[["Hobby"]]
```
```{r}
class(list1["Hobby"])
class(list1[["Hobby"]])
```

```{exercise}
Exercise 3.19        
Based on the data set of sale.tag, choose all correct answers from the following options A - F.
  A: sale.tag[[1]]
  B: sale.tag[[Spring]]
  C: sale.tag[["Spring"]]
  D: sale.tag["Spring"]
  E: sale.tag[1]
  F: sale.tag[Spring]
  
1, Which options return a vector.  Answer:___________ 
2, Which options return a list.    Answer:___________
3, Which options returns an error. Answer:___________
````

Solution:             
1, Answer: A, C             
2, Answer: D, E          
3, Answer: B, F    


```{Exercise}
Exercise 3. 20

Fill blanks in the function below so that for a given list, the function returns all locations of each element and sort the frequency of the elements in an increasing order. Then apply your function to the set of gender:
  F F M M M M I I F I F M I F M I I M I I F
Where F, M and I represent Female, Male and Infant respectively.

```

Solultion:

```{r}
my.fun <- function(f){
y <- unlist(f)                     
x <- list()                        
for (i in 1:length(y)){        
  c.name <- y[i]                
  x[[c.name]] <- c(x[[c.name]],i) 
}
freq <- sapply(x, length)
b1 <- sort(freq, decreasing = F)
lst <- list(x, b1)
return(lst)
}
```

```{r}
gender <- list( "F","F", "M", "M", "M", "M", "I", "I", "F", "I", "F", "M", "I", "F", "M", "I", "I", "M", "I", "I", "F")
my.fun(gender)
```

